---
title: "Introduction to R"
author: "Adam Johnson"
date: '`r format(Sys.Date(), "%m/%d/%Y")`'
output:
  word_document: default
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is a getting started guide for R, and contains concepts from articles I found helpful and introductory material from RStudio.  Read this if you've been away from R for awhile and need a quick refresher, or if you're just starting out and need an introduction.

# R and RStudio

R is a programming language commonly used for data analytics.  [RStudio](https://rstudio.com/) is a company which makes an IDE (also called RStudio) commonly used with the R language.  If you open the RStudio IDE (with the default layout), you'll see several windows arranged as follows:

* Upper Left
  * **Source**: .R, .Rmd, .Rpres, etc. files open here
    * *Source* executes a script file; makes functions available in the Console.
    * *Knit* compiles an R Markdown file.
* Lower Left
  * **Console**:  R interpreter with command line; Working directory shown at top
* Upper Right
  * **Environment**: generated data structures
    * *Broom* deletes existing sourced data.
  * **History**: past commands to run/copy
    * *To Console* pastes a command into the Console window.
    * *To Source* pastes a commmand into the Source window.
  * **Presentation**: displays .RPres slideshows
* Lower Right
  * **Files**: file browser keyed to working directory
    * *More --> Go To Working Directory* to view.
    * *More --> Set As Working Directory* to set.
  * **Plots**: generated graphs
  * **Packages**: package manager
    * *Install* installs package(s).
    * *Update* updates package(s).
  * **Viewer**: interactive data exploration
  * **Help**: extensive documentation (also in the Help menu)

After opening RStudio, remember to:

* Set the working directory.
* Source the R scripts.
* Load required libraries (perhaps you have a function to do this in your scripts?).
* Use Ctrl+Z = undo; Ctrl+Shift+Z = redo.

# File types

There are several types of R files.

## R Script

A *.R* file contains code in the R language.  You can organize the code into functions, create libraries and packages, etc., just as you'd expect.

## R Markdown

A *.Rmd* file, which stands for *R Markdown*, can contain sections of R code, and also sections of [Markdown](https://github.com/adam-p/markdown-here/wiki/Markdown-Cheatsheet) text.  These files can be compiled, or *knit*, to generate a report, which is a document containing any combination of formatted text, R code, and output (such as plots) from R code.  Sections of R code in a Rmd file can include parameters which specify what happens when a report is generated.  These are some important ones:

- **echo**:  include the source code in the generated report
- **eval**:  run the code
- **include**:  include code output (like plots) in report

The report generated from an R markdown file can be a PDF, Word document, or HTML file.  As you can imagine, this can be *really* handy for processing data, which can automatically be formatted into a report.  In the *INSERT NAME OF THE PART OF R STUDIO HERE*, click *Knit* to generate the report for this file, and you'll have a handy dandy getting started guide to read while you get oriented to the language and IDE.  For this file, *Knit to Word* has been tested, the other output types aren't guaranteed to work (yet!).

## R Presentation

A *.Rpres* file, which stands for *R Presentation*, will generate a PowerPoint-like output which is shown in the *Presentation* tab which opens in the top-right portion of the RStudio IDE.  In RStudio, go to *File --> New File --> R Presentation*.  A new presentation will be created, and it will contain samples of how to create slides, embed text, bullet points, links, code, and plots.  Click *Preview* to see the presentation output.

# Data Structures

The material in this section is based on a very helpful article called [Advanced R](http://adv-r.had.co.nz/Data-structures.html).

R’s base data structures can be organised by their dimensionality (1d, 2d, or nd) and whether they’re homogeneous (all contents must be of the same type) or heterogeneous (the contents can be of different types). This gives rise to the five data types most often used in data analysis:

|  |  Homogeneous  | Heterogeneous |
|--|---------------|---------------|
|1d| Atomic vector | List          |
|2d| Matrix        | Data frame    |
|nd| Array         |               |

Use str() to fetch the structure of an object.

## Vectors

Can be atomic vectors or lists, and have type, length, and attributes.
Atomic vectors are usually created with c(), short for combine:

```{r atomic_vectors, echo=TRUE, eval=FALSE}
dbl_var <- c(1, 2.5, 4.5)

# With the L suffix, you get an integer rather than a double
int_var <- c(1L, 6L, 10L)

# Use TRUE and FALSE (or T and F) to create logical vectors
log_var <- c(TRUE, FALSE, T, F)

chr_var <- c("these are", "some strings")
```

Atomic vectors are always flat, even if you nest c()’s:

```{r always_flat, echo=TRUE}
# Both lines have the same output:
c(1, c(2, c(3, 4)))

c(1, 2, 3, 4)

```

### Coercion

All elements of an atomic vector must be the same type, so when you attempt to combine different types they will be coerced to the most flexible type. Types from least to most flexible are: logical, integer, double, and character.

For example, combining a character and an integer yields a character:

```{r coercion, echo=TRUE}
str(c("a", 1))
```

When a logical vector is coerced to an integer or double, TRUE becomes 1 and FALSE becomes 0. This is very useful in conjunction with sum() and mean()

```{r conjunction, echo=TRUE}
x <- c(FALSE, FALSE, TRUE)
as.numeric(x)

# Total number of TRUEs
sum(x)

# Proportion that are TRUE
mean(x)
```

## Lists

Unlike atomic vectors because their elements can be any type, including other lists.  Created by list().

```{r lists, echo=TRUE}
x <- list(1:3, "a", c(TRUE, FALSE, TRUE), c(2.3, 5.9))
str(x)
```

Lists are sometimes called recursive vectors, because a list can contain other lists. This makes them fundamentally different from atomic vectors.

```{r recursive_vector, echo=TRUE}
x <- list(list(list(list())))
str(x)
is.recursive(x)
```

c() will combine several lists into one. If given a combination of atomic vectors and lists, c() will coerce the vectors to lists before combining them. Compare the results of list() and c():

```{r combine_lists, echo=TRUE}
x <- list(list(1, 2), c(3, 4))
y <- c(list(1, 2), c(3, 4))
str(x)
str(y)
```

## Attributes

All objects can have arbitrary additional attributes, used to store metadata about the object. Attributes can be thought of as a named list (with unique names). Attributes can be accessed individually with attr() or all at once (as a list) with attributes().  Most attributes are lost when modifying a vector, except for names, dimensions, and class.

```{r attributes, echo=TRUE}
y <- 1:10
attr(y, "my_attribute") <- "This is a vector"
attr(y, "my_attribute")

str(attributes(y))

structure(1:10, my_attribute = "This is a vector")
```

You can name a vector when creating it, modifying it, or copying it:

```{r naming_vectors_creation, echo=TRUE}
# Naming a vector during creation.
x <- c(a = 1, b = 2, c = 3)
```
```{r naming_vectors_modification, echo=TRUE}
# Naming a vector during modification.
x <- 1:3; names(x) <- c("a", "b", "c")
```
```{r naming_vectors_modification_alternate, echo=TRUE}
# Another way to do the same thing.
x <- 1:3; names(x)[[1]] <- c("a")
```
```{r naming_vectors_copying, echo=TRUE}
# Naming a vector by copying.
x <- setNames(1:3, c("a", "b", "c"))
```

Not all elements of a vector need to have a name. If some names are missing when you create the vector, the names will be set to an empty string for those elements. If you modify the vector in place by setting some, but not all variable names, names() will return NA (more specifically, NA_character_) for them. If all names are missing, names() will return NULL.

```{r vector_names, echo=TRUE}
y <- c(a = 1, 2, 3)
names(y)

v <- c(1, 2, 3)
names(v) <- c('a')
names(v)

z <- c(1, 2, 3)
names(z)
```

### Factors

One important use of attributes is to define factors. A factor is a vector that can contain only predefined values, and is used to store categorical data. Factors are built on top of integer vectors using two attributes: the class, “factor”, which makes them behave differently from regular integer vectors, and the levels, which defines the set of allowed values.

```{r echo=TRUE}
x <- factor(c("a", "b", "b", "a"))
x

class(x)
levels(x)

# You can't use values that are not in the levels
x[2] <- "c"
x

# NB: you can't combine factors
c(factor("a"), factor("b"))
```

Factors are useful when you know the possible values a variable may take, even if you don’t see all values in a given dataset. Using a factor instead of a character vector makes it obvious when some groups contain no observations:

```{r factor_example, echo=TRUE}
sex_char <- c("m", "m", "m")
sex_factor <- factor(sex_char, levels = c("m", "f"))

table(sex_char)

table(sex_factor)
```

## Matrices and Arrays

Adding a dim attribute to an atomic vector allows it to behave like a multi-dimensional array. A special case of the array is the matrix, which has two dimensions. Matrices are used commonly as part of the mathematical machinery of statistics. Arrays are much rarer.

Matrices and arrays are created with matrix() and array(), or by using the assignment form of dim():

```{r array_creation, echo=TRUE}
# Two scalar arguments to specify rows and columns
a <- matrix(1:6, ncol = 3, nrow = 2)

# One vector argument to describe all dimensions
b <- array(1:12, c(2, 3, 2))

# You can also modify an object in place by setting dim()
c <- 1:6
dim(c) <- c(3, 2)
c

dim(c) <- c(2, 3)
c
```

## Data Frames

A data frame is the most common way of storing data in R, and if used systematically makes data analysis easier. Under the hood, a data frame is a list of equal-length vectors. This makes it a 2-dimensional structure, so it shares properties of both the matrix and the list. This means that a data frame has names(), colnames(), and rownames(), although names() and colnames() are the same thing. The length() of a data frame is the length of the underlying list and so is the same as ncol(); nrow() gives the number of rows.

You create a data frame using data.frame(), which takes named vectors as input.  Beware data.frame()’s default behaviour which turns strings into factors.  Use stringsAsFactors = FALSE to suppress this behaviour:

```{r data_frame_creation, echo=TRUE}
df <- data.frame(x = 1:3, y = c("a", "b", "c"))
str(df)

df <- data.frame(x = 1:3, y = c("a", "b", "c"), stringsAsFactors = FALSE)
str(df)
```

You can combine data frames using cbind() and rbind().  When combining column-wise, the number of rows must match, but row names are ignored. When combining row-wise, both the number and names of columns must match.

```{r data_frame_combination, echo=TRUE}
# Column bind
cbind(df, data.frame(z = 3:1))

# Row bind
rbind(df, data.frame(x = 10, y = "z"))
```

It’s a common mistake to try and create a data frame by cbind()ing vectors together. This doesn’t work because cbind() will create a matrix unless one of the arguments is already a data frame. Instead use data.frame() directly:

```{r, echo=TRUE}
bad <- data.frame(cbind(a = 1:2, b = c("a", "b")))
str(bad)

good <- data.frame(a = 1:2, b = c("a", "b"), stringsAsFactors = FALSE)
str(good)
```

# Regression

```{r, regression, echo=TRUE}
# Create a data set that we can perform regression on.
A <- structure(list(Time = c(0, 1, 2, 4, 6, 8, 9, 10, 11, 12, 13, 14, 15,
                             16, 18, 19, 20, 21, 22, 24, 25, 26, 27, 28, 29, 30), 
                    Counts = c(126.6, 101.8, 71.6, 101.6, 68.1, 62.9, 45.5, 41.9,
                               46.3, 34.1, 38.2, 41.7, 24.7, 41.5, 36.6, 19.6, 22.8,
                               29.6, 23.5, 15.3, 13.4, 26.8, 9.8, 18.8, 25.9, 19.3)),
               .Names = c("Time", "Counts"),
               row.names = c(1L, 2L, 3L, 5L, 7L, 9L, 10L, 11L, 12L, 13L, 14L, 15L, 16L,
                             17L, 19L, 20L, 21L, 22L, 23L, 25L, 26L, 27L, 28L, 29L, 30L, 31L),
               class = "data.frame")
attach(A)
names(A)

# Perform linear regression.
linearFit <-lm(Counts ~ Time)

# See a summary, and then plot it.
summary(linearFit)
plot(Time, Counts, pch=16, ylab = "Counts ", col = "red" )
abline(linearFit, col = "blue")

# Perform a quadratic fit.
Time2 <- Time^2
quadraticFit <-lm(Counts ~ Time + Time2)

# See a summary, and then plot it on the same graph.
summary(quadraticFit)
timevalues <- seq(0, 30, 0.1)
predictedcounts <- predict(quadraticFit, list(Time=timevalues, Time2=timevalues^2))
lines(timevalues, predictedcounts, col = "darkgreen")

```

# Other resources:
* [RStudio Cheat Sheets](https://rstudio.com/resources/cheatsheets/)
* [Using R in C#](https://coders-corner.net/2015/11/01/use-r-in-c/)
* [Intro to Git & GitHub for the R User](https://github.com/bcgov/bcgov-data-science-resources/wiki/Tutorial:-Intro-to-Git-&-GitHub-for-the-R-User)
